$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-04-30 paper/paper.tex
Comment on pair synonyms
Align typing rules

% Fields
\newcommand{\op}[3]{#1 \; \field{\coloneqq} \; #2 \; \field{\cdot} \; #3}
\newcommand{\zero}{\field{0\cdot}}
\newcommand{\one}{\field{1\cdot}}
\newcommand{\li}{\field{\ell_i}}
\newcommand{\lo}{\field{\ell_o}}
\newcommand{\lz}{\field{\ell_{\o}}}
\newcommand{\lio}{\field{\ell_{\#}}}

@@

\begin{note}
  We will often work with input and output multiplicities as pairs.
  We use the notation $\Carrier^2$ to stand for a $\Carrier \times \Carrier$ tuple, and $\opsquared{x}{y}{z}$ to stand for a monoidal operation on pairs of multiplicities -- where all properties are trivially lifted.
\end{note}

@@

  Scope restriction expects three arguments: the type $t$ of data being transmitted; the input and output multiplicities $x$ of the data being transmitted; and the multiplicity $y$ assigned to the channel itself.
  The same multiplicity $y$ is used as the input multiplicity and as the output multiplicity of the channel, so these are balanced.
  The typing derivation of the continuation process $P$ is provided with this channel with multiplicity $(y \comma y)$.
  Process $P$ must then completely exhaust this channel, and to that end we substitute $(y , y)$ in the input context with $\ell_{\o}$ in the output context.

  \todo{comment on synonyms $\ell_i$ etc}

  Receiving requires a proof $chan_i$ that selects a channel at index $i$ with multiplicity $\ell_i$ available, over which data with type $t$ and multiplicity $x$ is going to be sent.
  Note that this index $i$ is then used in the syntax of the typed process.
  After consuming $\ell_i$ from this channel, we use the leftover context $\Xi$ to type the continuation.
  The continuation is also provided with the received element --- of type $t$ and multiplicity $x$ at index $\constr{zero}$.
  This multiplicity $x$ must be completely exhausted by the continuation process.

  Similar to input, output requires a proof $chan_i$ that selects a channel at index $i$ with multiplicity $\ell_o$ available, over which data with type $t$ and multiplicity $x$ is going to be sent.
  Note that this index $i$ is then used in the syntax of the typed process.
  After consuming $\ell_o$ from this channel, we use the leftover context $\Delta$ to type the continuation.
  The continuation must select the element at index $j$ that is being sent.
  This element must have type $t$ and at least multiplicity $x$ --- as per the channel's type.
  The result of consuming these multiplicity $x$ is the leftover context $\Xi$, which is then used to type the continuation process.
  Note that both indices $i$ and $j$ are used in the syntax of the typed process.

  Finally, parallel composition is only concerned with usage contexts.
  The leftovers of the left-hand process are used to type the right-hand process.

@@

    \inferrule
        {\stacked{
            chan_i : \contains{\gamma}{\Gamma}{i}{\channel{t}{x}}{\li}{\Xi} \\\\
            cont \hspace{0.6em} : \types{\gamma \comma t}{\Xi \comma x}{P}{\Theta \comma \lz}}}
        {\constr{recv} \; chan_i \; cont : \types{\gamma}{\Gamma}{\recv{i}{P}}{\Theta}}

    \inferrule
        {\stacked{
            chan_i : \contains{\gamma}{\Gamma}{i}{\channel{t}{x}}{\lo}{\Delta} \\\\
            loc_j \hspace{0.8em} : \contains{\gamma}{\Delta}{j}{t}{x\hspace{4em}}{\Xi} \\\\
            cont \hspace{0.6em} : \types{\gamma}{\Xi}{P\hspace{6em}}{\Theta}}}
        {\constr{send} \; chan_i \; loc_j \; cont : \types{\gamma}{\Gamma}{\send{i}{j}P}{\Theta}}

    \inferrule
    {l : \types{\gamma}{\Gamma}{P}{\Delta} \\\\
     r : \types{\gamma}{\Delta}{Q}{\Xi}}
    {\constr{comp} \; l \; r : \types{\gamma}{\Gamma}{\comp{P}{Q}}{\Xi}}
  \end{mathpar}
  \caption{Leftover typing for a resource-aware typing system.}
  \label{fig:types}
  \end{figure}
  \todo{align typing rules to make them more readable}
\end{nidefinition}

\subsubsection{Example derivations}

@@

\begin{proof}
  By induction on \textsc{Equals}.
  The case $\constr{struct}$ and its symmetric one under $\constr{sym}$ are the nontrivial ones.
  For those, we proceed by induction on \textsc{StructCong}:
  \begin{itemize}
    \item
      The case $\constr{comp-assoc}$ is trivial because the encoding with leftover typing of the typing rules naturally preserves associativity.
    \item
      The $\constr{comp-sym}$ case for a process $\comp{P}{Q}$ is however more complicated: $P$ must use framing (Theorem \ref{thm:framing}) to shift its output context to $Q$'s; $Q$ must use framing to shift its input context to $P$'s.
      The usage context on which $Q$ and $P$ coincide in $\comp{Q}{P}$ will no longer be the same.
    \item
      The case $\constr{comp-end}$ is trivial because in the typing rule for process $\PO$ the input and output contexts are one and the same.
    \item
      In the $\constr{scope-end}$ case, the typing rules $\constr{chan}$ and $\constr{end}$ work together to constraint the multiplicity of the newly create channel to $\ell_{\o}$, making the proof trivial.
      In the opposite direction, we will need to instantiate the newly created channel to a concrete type and multiplicities.
      For this, we will use type $\unit$ and multiplicities $\ell_{\o}$.
    \item
      In the $\constr{scope-ext}$ case for a process $\new \comp{P}{Q}$, we will make use of Lemma $\ref{lm:types-unused}$ to show that as the typing derivation on $P$ will preserved the multiplicity of the recently introduced variable unchanged.
      We then use strengthening (Theorem \ref{thm:strengthening}) to delete the recently introduced variable from $P$, and then compose with the typing derivation of $Q$ under $\constr{chan}$.
      In the reverse direction, we will weaken $P$ so that it can accept an additional variable.
      We will then use Lemma \ref{lm:lower-lift} to show that lifting a lowered $P$ leaves $P$ unchanged.
    \item
      Finally, for the $\constr{scope-comm}$ rule, we will use swapping as shown in Theorem \ref{thm:swapping}.
      In the reverse direction, we will use swapping and then use Lemma \ref{lm:swap-swap} to show that swapping indices twice leaves $P$ unchanged.
  \end{itemize}
\end{proof}
\todo{make sure constructor names in code coincide}

@@

\begin{nitheorem}
  \label{thm:substitution}
  Let $P$ be a process where the most recent variable has all its $m$ multiplities exhausted, such that $\types{\gamma \comma t}{\Gamma \comma m}{P}{\Psi \comma \ell_{\o}}$.
  Let $j$ be the index of the data with $m$ multiplicities that is being received after $P$ is done, so that $\contains{\gamma}{\Psi}{j}{t}{m}{\Xi}$.
  Then, we can substitute variable $\constr{zero}$ in $P$ with $\suc j$.
  As a result, the multiplicities at $\constr{zero}$ will be preserved and those at $j$ will be consumed instead, resulting in $\types{\gamma \comma t}{\Gamma \comma m}{\subst{P}{\suc j}{\constr{zero}}}{\Xi \comma m}$.
\end{nitheorem}
\begin{proof}
  We need to proceed by induction on $P$, and to do so means going under scope restriction.
  We use the generalised version in Theorem \ref{thm:subst-generalization} so that the variable that is being substituted is not $\constr{zero}$ but any variable $i$.
  \todo{explain arguments}
\end{proof}

@@

  Let the variable that is being substituted be $\contains{\gamma}{\Gamma_i}{i}{t}{m}{\Gamma}$ and the variable that it is being substituted with $\contains{\gamma}{\Gamma_j}{j}{t}{m}{\Gamma}$.
  As shown in Figure \ref{fig:subst}, the process that is being substituted will pivot from the input usage context $\Gamma_i$ to the input usage context $\Gamma_j$.
  Let there be two \emph{adjustment} variables $\contains{\gamma}{\Psi_i}{i}{t}{n}{\Psi}$ and $\contains{\gamma}{\Psi_j}{j}{t}{n}{\Psi}$.
  The process that is being substituted will pivot from the output usage context $\Psi_i$ to the output usage context $\Psi_j$.
  As shown in Figure \ref{fig:subst-alternative}, without this \emph{adjustment} composing the substitution of subprocesses is not at all trivial.
  Assume that $P$ uses all $m$ multiplicities at index $i$.
  To model this, require $\opctx{\Gamma}{\Delta}{\Psi}$ for some $\Delta$ where index $i$ is $\ell_{\o}$.
  The process that is typed such that $\types{\gamma}{\Gamma_i}{P}{\Psi_i}$ will be typed $\types{\gamma}{\Gamma_j}{\subst{P}{j}{i}}{\Psi_j}$ after substitution.
\end{nitheorem}

@@

      \draw[-]  (gamma-m) -- (delta-m.center);
      \draw[->] (delta-m.center) -- node[align=center,below] {$\Delta$\\$\Delta_i = \ell_{\o}$}(psi-m);

@@

\begin{proof}
  By induction on the derivation $\types{\gamma}{\Gamma_i}{P}{\Psi_i}$.
  \begin{itemize}
    \item
      From constructor $\constr{end}$ we derive $\Gamma_i \equiv \Psi_i$.
      From $\Delta_i \equiv \ell_{\o}$ follows that $m \equiv n$.
      Therefore $\Gamma_j \equiv \Psi_j$ and $\constr{end}$ can be applied.

@@

    \item
      For constructor $\constr{recv}$ we split $\Delta$ so that we can proceed inductively on the continuation.
      We first observe that given that there exists an arrow from $\Gamma_i$ to $\Psi_i$ and $\Delta$ at index $i$ is $\ell_{\o}$, for any $m$ and $n$, there is some $\delta$ such that $\opsquared{m}{\delta}{n}$.
 l     \begin{itemize}
        \item
          If the input occurs on the variable that is being substituted, we derive $\opsquared{m}{\ell_i}{l}$ for some $l$ and construct an arrow $\containsusage{\Xi_i}{i}{l}{\Gamma}$ (used for the inductive call).
          Similarly, we construct the arrows $\containsusage{\Gamma_j}{j}{\ell_i}{\Xi_j}$ (used as the new channel on which input occurs) and $\containsusage{\Xi_j}{j}{l}{\Gamma}$ (used for the inductive call) for some $\Xi_j$.
        \item
          If the input occurs on some variable $x$ other than the one that is being substituted, we construct the arrows $\containsusage{\Xi_i}{i}{m}{\Theta}$ (used for the inductive call) and $containsusage{\Gamma}{x}{\ell_i}{\Theta}$ for some $\Theta$.
          We can then find $\Xi_j$ and construct the two arrows $\containsusage{\Gamma_j}{x}{\ell_i}{\Xi_j}$ (used as the new channel on which input occurs) and $\containsusage{Xi_j}{j}{m}{\Theta}$ (used for the inductive call).
          From the fact that there exists a composition of arrows from $\Xi_i$ to $\Psi$, we can conclude that $\Theta$ is splitting $\Delta$ such that $\opctx{\Gamma}{\Delta_1}{\Theta}$ and $\opctx{\Theta}{\Delta_2}{\Psi}$.
          As $\ell_{\o}$ is a minimal element, if $\Delta$ is $\ell_{\o}$ at index $i$, then it must be $\ell_{\o}$ at index $i$ for $\Delta_1$ and $\Delta_2$ as well.
      \end{itemize}

@@

    \item
      For $\constr{comp}$ we first find a $\delta$, $\Theta$, $\Delta_1$ and $\Delta_2$ such that $\containsusage{\Xi_i}{i}{\delta}{\Theta}$ and $\opctx{\Gamma}{\Delta_1}{\Theta}$ and $\opctx{\Theta}{\Delta_2}{\Psi}$.
      Given that $\Delta$ is $\ell_{\o}$ at index $i$, we know that $\Delta_1$ and $\Delta_2$ are so too.
      We observe that $\opsquared{m}{\delta}{\omega}$ where $\omega$ is what subprocess $P$ consumes at index $i$.
      We use this to construct an arrow $\containsusage{\Xi_j}{j}{\delta}{\Theta}$, for some $\Xi_j$.
      We use these constructions to make two inductive calls (on the derivation of $P$ and $Q$) and compose their result.
  \end{itemize}  
\end{proof}

@@

\begin{nitheorem}
  Let $P \reduce{c} Q$.
  If $c$ is $\constr{internal}$ to the process, the context must not be modified and thus $\Gamma' \equiv \Gamma$.
  If $c$ is $\constr{external}\; i$, the input usage context at index $i$ must be decremented, and therefore $\containsusage{\Gamma'}{i}{\ell_\#}{\Gamma}$.
  Then the typing derivation $\types{\gamma}{\Gamma'}{P}{\Xi}$ can be transformed into $\types{\gamma}{\Gamma}{Q}{\Xi}$.
\end{nitheorem}

@@

    \begin{figure}[h]
      \setlength{\mathindent}{0pt}
      \begin{subfigure}{.3\textwidth}
        \begin{alignat*}{2}
          \containsusage{&\Gamma'}{i}{\ell_\# &&}{\Gamma} \\
          \contains{\gamma}{&\Gamma'}{i}{\channel{t}{m}}{\ell_i &&}{\Psi} \\
          \types{\gamma \comma t}{&\Psi \comma m}{P &&}{\Delta \comma \ell_{\o}} \\
          \contains{\gamma}{&\Delta }{i}{\channel{t}{m}}{\ell_o &&}{\Theta} \\
          \contains{\gamma}{&\Theta}{j}{t}{m &&}{\Xi}
        \end{alignat*}
        \caption{Initial assumptions.}
        \label{fig:subject-reduction:a}
      \end{subfigure}
      % Do not leave empty space
      \begin{subfigure}{.5\textwidth}
        \begin{alignat*}{2}
          \types{\gamma \comma t}{& \Gamma \comma m}{P &&}{\Theta \comma \ell_{\o}} \\
          \contains{\gamma}{& \Theta}{j}{t}{m &&}{\Xi}
        \end{alignat*}
        \caption{Rearrange $\ell_o$ witness, and merge with $\ell_i$ witness into $\ell_\#$.}
      \end{subfigure}
      % Do not leave empty space
      \begin{subfigure}{.5\textwidth}
        \begin{align*}
          \types{\gamma \comma t}{\Gamma \comma m}{\subst{P}{\suc j}{\constr{zero}}}{\Xi \comma m}
        \end{align*}
        \caption{Substitution.}
      \end{subfigure}
      % Do not leave empty space
      \begin{subfigure}{.5\textwidth}
        \begin{align*}
          \types{\gamma \comma t}{\Gamma \comma m}{lower_0 \; (\subst{P}{\suc j}{\constr{zero}})}{\Xi \comma m}
        \end{align*}
        \caption{Strengthening.}
      \end{subfigure}
      \caption{Proof steps for the $\constr{comm}$ case in subject reduction.}
    \end{figure}

@@

    \item
    Reduction under scope restriction case splits on the channel $c$ on which communication occurs.
    If the channel is $\constr{internal}$ to the process subject reduction proceeds inductively.
    If the channel is $\constr{external}\; \constr{zero}$ -- i.e. the channel that is being introduced by the scope restriction -- we use the Lemma \ref{lm:comm-capable} to subtract $\ell_\#$ from the channel's multiplicity first, then apply the induction hypothesis.
    If the channel is $\constr{external}\; (\suc i)$ subject reduction proceeds inductively.

@@

\begin{nilemma}
  \label{lm:comm-capable}
  Every input usage context $\Gamma$ on a well-typed process $\types{\gamma}{\Gamma}{P}{\Delta}$ that reduces such that $P \reduce{\constr{external} \; i} Q$ has a multiplicity of at least $\ell_\#$ at index $i$.
\end{nilemma}
\begin{proof}
  By induction on the reduction derivation $P \reduce{\constr{external \; i}}Q$.
\end{proof}

>>>>

% Fields
\newcommand{\op}[3]{#1 \; \field{\coloneqq} \; #2 \; \field{\cdot} \; #3}
\newcommand{\zero}{\field{0\cdot}}
\newcommand{\one}{\field{1\cdot}}
\newcommand{\li}{\ell_i}
\newcommand{\lo}{\ell_o}
\newcommand{\lz}{\ell_{\o}}
\newcommand{\lio}{\ell_{\#}}

@@

\begin{note}
  We will often work with input and output multiplicities as pairs.
  We use the notation $\Carrier^2$ to stand for a $\Carrier \times \Carrier$ tuple, and $\opsquared{x}{y}{z}$ to stand for a monoidal operation on pairs of multiplicities -- where all properties are trivially lifted.
  Henceworth, we use $\lz$ to denote the pair $(0 \comma 0)$, $\li$ for the pair $(1 \comma 0)$, $\lo$ for $(0 \comma 1)$ and $\lio$ for $(1 \comma 1)$.
\end{note}

@@

  Scope restriction expects three arguments: the type $t$ of data being transmitted; the input and output multiplicities $x$ of the data being transmitted; and the multiplicity $y$ assigned to the channel itself.
  The same multiplicity $y$ is used as the input multiplicity and as the output multiplicity of the channel, so these are balanced.
  The typing derivation of the continuation process $P$ is provided with this channel with multiplicity $(y \comma y)$.
  Process $P$ must then completely exhaust this channel, and to that end we substitute $(y , y)$ in the input context with $\lz$ in the output context.

  Receiving requires a proof $chan_i$ that selects a channel at index $i$ with multiplicity $\li$ available, over which data with type $t$ and multiplicity $x$ is going to be sent.
  Note that this index $i$ is then used in the syntax of the typed process.
  After consuming $\li$ from this channel, we use the leftover context $\Xi$ to type the continuation.
  The continuation is also provided with the received element --- of type $t$ and multiplicity $x$ at index $\constr{zero}$.
  This multiplicity $x$ must be completely exhausted by the continuation process.

  Similar to input, output requires a proof $chan_i$ that selects a channel at index $i$ with multiplicity $\lo$ available, over which data with type $t$ and multiplicity $x$ is going to be sent.
  Note that this index $i$ is then used in the syntax of the typed process.
  After consuming $\lo$ from this channel, we use the leftover context $\Delta$ to type the continuation.
  The continuation must select the element at index $j$ that is being sent.
  This element must have type $t$ and at least multiplicity $x$ --- as per the channel's type.
  The result of consuming these multiplicity $x$ is the leftover context $\Xi$, which is then used to type the continuation process.
  Note that both indices $i$ and $j$ are used in the syntax of the typed process.

  Parallel composition is only concerned with usage contexts.
  It uses the leftovers of the left-hand process to type the right-hand process.

@@

    \inferrule
        {\stacked{
            chan_i : \contains{\gamma \hspace{1.2em}}{\Gamma \hspace{1.3em}}{i}{\channel{t}{x}}{\li}{\Xi} \\\\
            cont \hspace{0.6em} : \types{\gamma \comma t}{\Xi \comma x}{P \hspace{5.6em}}{\Theta \comma \lz}}}
        {\constr{recv} \; chan_i \; cont : \types{\gamma}{\Gamma}{\recv{i}{P}}{\Theta}}

    \inferrule
        {\stacked{
            chan_i : \contains{\gamma}{\Gamma \hspace{0.2em}}{i}{\channel{t}{x}}{\lo}{\Delta} \\\\
            loc_j \hspace{0.8em} : \contains{\gamma}{\Delta}{j}{t \hspace{3.7em}}{x \hspace{0.3em}}{\Xi} \\\\
            cont \hspace{0.6em} : \types{\gamma}{\Xi \hspace{0.3em}}{P\hspace{5.8em}}{\Theta}}}
        {\constr{send} \; chan_i \; loc_j \; cont : \types{\gamma}{\Gamma}{\send{i}{j}P}{\Theta}}

    \inferrule
    {l \hspace{0.3em} : \types{\gamma}{\Gamma \hspace{0.2em}}{P}{\Delta} \\\\
     r : \types{\gamma}{\Delta}{Q}{\Xi}}
    {\constr{comp} \; l \; r : \types{\gamma}{\Gamma}{\comp{P}{Q}}{\Xi}}
  \end{mathpar}
  \caption{Leftover typing for a resource-aware typing system.}
  \label{fig:types}
  \end{figure}
\end{nidefinition}

\subsubsection{Example derivations}
\label{example-derivations}

@@

\begin{proof}
  By induction on \textsc{Equals}.
  The case $\constr{struct}$ and its symmetric one under $\constr{sym}$ are the nontrivial ones.
  For those, we proceed by induction on \textsc{StructCong}:
  \begin{itemize}
    \item
      The case $\constr{comp-assoc}$ is trivial because the encoding with leftover typing of the typing rules naturally preserves associativity.
    \item
      The $\constr{comp-sym}$ case for a process $\comp{P}{Q}$ is however more complicated: $P$ must use framing (Theorem \ref{thm:framing}) to shift its output context to $Q$'s; $Q$ must use framing to shift its input context to $P$'s.
      The usage context on which $Q$ and $P$ coincide in $\comp{Q}{P}$ will no longer be the same.
    \item
      The case $\constr{comp-end}$ is trivial because in the typing rule for process $\PO$ the input and output contexts are one and the same.
    \item
      In the $\constr{scope-end}$ case, the typing rules $\constr{chan}$ and $\constr{end}$ work together to constraint the multiplicity of the newly create channel to $\lz$, making the proof trivial.
      In the opposite direction, we will need to instantiate the newly created channel to a concrete type and multiplicities.
      For this, we will use type $\unit$ and multiplicities $\lz$.
    \item
      In the $\constr{scope-ext}$ case for a process $\new \comp{P}{Q}$, we will make use of Lemma $\ref{lm:types-unused}$ to show that as the typing derivation on $P$ will preserved the multiplicity of the recently introduced variable unchanged.
      We then use strengthening (Theorem \ref{thm:strengthening}) to delete the recently introduced variable from $P$, and then compose with the typing derivation of $Q$ under $\constr{chan}$.
      In the reverse direction, we will weaken $P$ so that it can accept an additional variable.
      We will then use Lemma \ref{lm:lower-lift} to show that lifting a lowered $P$ leaves $P$ unchanged.
    \item
      Finally, for the $\constr{scope-comm}$ rule, we will use swapping as shown in Theorem \ref{thm:swapping}.
      In the reverse direction, we will use swapping and then use Lemma \ref{lm:swap-swap} to show that swapping indices twice leaves $P$ unchanged.
  \end{itemize}
\end{proof}
\todo{make sure constructor names in code coincide}

@@

\begin{nitheorem}
  \label{thm:substitution}
  Let $P$ be a process where the most recent variable has all its $m$ multiplities exhausted, such that $\types{\gamma \comma t}{\Gamma \comma m}{P}{\Psi \comma \lz}$.
  Let $j$ be the index of the data with $m$ multiplicities that is being received after $P$ is done, so that $\contains{\gamma}{\Psi}{j}{t}{m}{\Xi}$.
  Then, we can substitute variable $\constr{zero}$ in $P$ with $\suc j$.
  As a result, the multiplicities at $\constr{zero}$ will be preserved and those at $j$ will be consumed instead, resulting in $\types{\gamma \comma t}{\Gamma \comma m}{\subst{P}{\suc j}{\constr{zero}}}{\Xi \comma m}$.
\end{nitheorem}
\begin{proof}
  We need to proceed by induction on $P$, and to do so means going under scope restriction.
  We use the generalised version in Theorem \ref{thm:subst-generalization} so that the variable that is being substituted is not $\constr{zero}$ but any variable $i$.
  \todo{explain arguments}
\end{proof}

@@

  Let the variable that is being substituted be $\contains{\gamma}{\Gamma_i}{i}{t}{m}{\Gamma}$ and the variable that it is being substituted with $\contains{\gamma}{\Gamma_j}{j}{t}{m}{\Gamma}$.
  As shown in Figure \ref{fig:subst}, the process that is being substituted will pivot from the input usage context $\Gamma_i$ to the input usage context $\Gamma_j$.
  Let there be two \emph{adjustment} variables $\contains{\gamma}{\Psi_i}{i}{t}{n}{\Psi}$ and $\contains{\gamma}{\Psi_j}{j}{t}{n}{\Psi}$.
  The process that is being substituted will pivot from the output usage context $\Psi_i$ to the output usage context $\Psi_j$.
  As shown in Figure \ref{fig:subst-alternative}, without this \emph{adjustment} composing the substitution of subprocesses is not at all trivial.
  Assume that $P$ uses all $m$ multiplicities at index $i$.
  To model this, require $\opctx{\Gamma}{\Delta}{\Psi}$ for some $\Delta$ where index $i$ is $\lz$.
  The process that is typed such that $\types{\gamma}{\Gamma_i}{P}{\Psi_i}$ will be typed $\types{\gamma}{\Gamma_j}{\subst{P}{j}{i}}{\Psi_j}$ after substitution.
\end{nitheorem}

@@

      \draw[-]  (gamma-m) -- (delta-m.center);
      \draw[->] (delta-m.center) -- node[align=center,below] {$\Delta$\\$\Delta_i = \lz$}(psi-m);

@@

\begin{proof}
  By induction on the derivation $\types{\gamma}{\Gamma_i}{P}{\Psi_i}$.
  \begin{itemize}
    \item
      From constructor $\constr{end}$ we derive $\Gamma_i \equiv \Psi_i$.
      From $\Delta_i \equiv \lz$ follows that $m \equiv n$.
      Therefore $\Gamma_j \equiv \Psi_j$ and $\constr{end}$ can be applied.

@@

    \item
      For constructor $\constr{recv}$ we split $\Delta$ so that we can proceed inductively on the continuation.
      We first observe that given that there exists an arrow from $\Gamma_i$ to $\Psi_i$ and $\Delta$ at index $i$ is $\lz$, for any $m$ and $n$, there is some $\delta$ such that $\opsquared{m}{\delta}{n}$.
 l     \begin{itemize}
        \item
          If the input occurs on the variable that is being substituted, we derive $\opsquared{m}{\li}{l}$ for some $l$ and construct an arrow $\containsusage{\Xi_i}{i}{l}{\Gamma}$ (used for the inductive call).
          Similarly, we construct the arrows $\containsusage{\Gamma_j}{j}{\li}{\Xi_j}$ (used as the new channel on which input occurs) and $\containsusage{\Xi_j}{j}{l}{\Gamma}$ (used for the inductive call) for some $\Xi_j$.
        \item
          If the input occurs on some variable $x$ other than the one that is being substituted, we construct the arrows $\containsusage{\Xi_i}{i}{m}{\Theta}$ (used for the inductive call) and $containsusage{\Gamma}{x}{\li}{\Theta}$ for some $\Theta$.
          We can then find $\Xi_j$ and construct the two arrows $\containsusage{\Gamma_j}{x}{\li}{\Xi_j}$ (used as the new channel on which input occurs) and $\containsusage{Xi_j}{j}{m}{\Theta}$ (used for the inductive call).
          From the fact that there exists a composition of arrows from $\Xi_i$ to $\Psi$, we can conclude that $\Theta$ is splitting $\Delta$ such that $\opctx{\Gamma}{\Delta_1}{\Theta}$ and $\opctx{\Theta}{\Delta_2}{\Psi}$.
          As $\lz$ is a minimal element, if $\Delta$ is $\lz$ at index $i$, then it must be $\lz$ at index $i$ for $\Delta_1$ and $\Delta_2$ as well.
      \end{itemize}

@@

    \item
      For $\constr{comp}$ we first find a $\delta$, $\Theta$, $\Delta_1$ and $\Delta_2$ such that $\containsusage{\Xi_i}{i}{\delta}{\Theta}$ and $\opctx{\Gamma}{\Delta_1}{\Theta}$ and $\opctx{\Theta}{\Delta_2}{\Psi}$.
      Given that $\Delta$ is $\lz$ at index $i$, we know that $\Delta_1$ and $\Delta_2$ are so too.
      We observe that $\opsquared{m}{\delta}{\omega}$ where $\omega$ is what subprocess $P$ consumes at index $i$.
      We use this to construct an arrow $\containsusage{\Xi_j}{j}{\delta}{\Theta}$, for some $\Xi_j$.
      We use these constructions to make two inductive calls (on the derivation of $P$ and $Q$) and compose their result.
  \end{itemize}  
\end{proof}

@@

\begin{nitheorem}
  Let $P \reduce{c} Q$.
  If $c$ is $\constr{internal}$ to the process, the context must not be modified and thus $\Gamma' \equiv \Gamma$.
  If $c$ is $\constr{external}\; i$, the input usage context at index $i$ must be decremented, and therefore $\containsusage{\Gamma'}{i}{\lio}{\Gamma}$.
  Then the typing derivation $\types{\gamma}{\Gamma'}{P}{\Xi}$ can be transformed into $\types{\gamma}{\Gamma}{Q}{\Xi}$.
\end{nitheorem}

@@

    \begin{figure}[h]
      \setlength{\mathindent}{0pt}
      \begin{subfigure}{.3\textwidth}
        \begin{alignat*}{2}
          \containsusage{&\Gamma'}{i}{\lio &&}{\Gamma} \\
          \contains{\gamma}{&\Gamma'}{i}{\channel{t}{m}}{\li &&}{\Psi} \\
          \types{\gamma \comma t}{&\Psi \comma m}{P &&}{\Delta \comma \lz} \\
          \contains{\gamma}{&\Delta }{i}{\channel{t}{m}}{\lo &&}{\Theta} \\
          \contains{\gamma}{&\Theta}{j}{t}{m &&}{\Xi}
        \end{alignat*}
        \caption{Initial assumptions.}
        \label{fig:subject-reduction:a}
      \end{subfigure}
      % Do not leave empty space
      \begin{subfigure}{.5\textwidth}
        \begin{alignat*}{2}
          \types{\gamma \comma t}{& \Gamma \comma m}{P &&}{\Theta \comma \lz} \\
          \contains{\gamma}{& \Theta}{j}{t}{m &&}{\Xi}
        \end{alignat*}
        \caption{Rearrange $\lo$ witness, and merge with $\li$ witness into $\lio$.}
      \end{subfigure}
      % Do not leave empty space
      \begin{subfigure}{.5\textwidth}
        \begin{align*}
          \types{\gamma \comma t}{\Gamma \comma m}{\subst{P}{\suc j}{\constr{zero}}}{\Xi \comma m}
        \end{align*}
        \caption{Substitution.}
      \end{subfigure}
      % Do not leave empty space
      \begin{subfigure}{.5\textwidth}
        \begin{align*}
          \types{\gamma \comma t}{\Gamma \comma m}{lower_0 \; (\subst{P}{\suc j}{\constr{zero}})}{\Xi \comma m}
        \end{align*}
        \caption{Strengthening.}
      \end{subfigure}
      \caption{Proof steps for the $\constr{comm}$ case in subject reduction.}
    \end{figure}

@@

    \item
    Reduction under scope restriction case splits on the channel $c$ on which communication occurs.
    If the channel is $\constr{internal}$ to the process subject reduction proceeds inductively.
    If the channel is $\constr{external}\; \constr{zero}$ -- i.e. the channel that is being introduced by the scope restriction -- we use the Lemma \ref{lm:comm-capable} to subtract $\lio$ from the channel's multiplicity first, then apply the induction hypothesis.
    If the channel is $\constr{external}\; (\suc i)$ subject reduction proceeds inductively.

@@

\begin{nilemma}
  \label{lm:comm-capable}
  Every input usage context $\Gamma$ on a well-typed process $\types{\gamma}{\Gamma}{P}{\Delta}$ that reduces such that $P \reduce{\constr{external} \; i} Q$ has a multiplicity of at least $\lio$ at index $i$.
\end{nilemma}
\begin{proof}
  By induction on the reduction derivation $P \reduce{\constr{external \; i}}Q$.
\end{proof}
